---
title: "Getting started with spinet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with spinet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r}
library(spinet)
set.seed(1)
```

## 1. The problem this package exists to solve

Suppose you have habitat suitability surfaces for a set of plants and a set of
pollinators, present and future, and you want to know how climate change will
reshape the interaction network. The obvious approach is to build a local
network in each grid cell from species co-occurrence and compare the two time
slices.

That approach cannot work, and it is worth seeing why before using anything
else in the package.

```{r}
P <- matrix(runif(500 * 12), 500, 12)     # 12 plants,     500 cells
A <- matrix(runif(500 * 10), 500, 10)     # 10 pollinators

f_naive <- si_overlap(P, A, metaweb = NULL, method = "product", check = FALSE)
si_degenerate(f_naive, warn = FALSE)[, 1:2]
```

The warning is not conservative. Every local matrix is
$W_k = p_k \otimes a_k$, which has rank one:

```{r}
w <- si_local(f_naive, 1, drop = FALSE)
c(rank        = qr(w)$rank,
  connectance = mean(w > 0),
  NODF        = si_nodf(w > 0),
  H2prime     = si_h2prime(w))
```

Connectance is exactly 1, nestedness exactly 0, and $H_2'$ exactly 0 — because
the maximum-entropy matrix with those marginals *is* the outer product of the
marginals. Nothing about network structure can vary across space, and nothing
can change between the present and the future. A map of any of these metrics is
a map of species richness wearing a disguise.

## 2. Adding a constraint layer

The fix is to record which pairs of species can actually interact. That is the
metaweb, or more generally a set of forbidden-link rules.

```{r}
mw <- si_simulate_metaweb(12, 10, connectance = 0.25, seed = 1)
mw

f <- si_overlap(P, A, mw, method = "product", floor = 0.05)
f
```

Now the metrics carry information:

```{r}
m <- si_metrics(f, what = c("links", "connectance", "NODF", "H2prime"))
summary(m)
```

Note the `floor` argument. Correlative suitability surfaces are strictly
positive almost everywhere, so with `floor = 0` the support of `W` equals the
metaweb in every cell and every presence-based metric returns the metaweb
itself. `si_prune()` sets a data-driven floor if you prefer not to choose one.

## 3. Building forbidden links from biology

Rules multiply, so a link survives only if every rule permits it.

```{r}
ph <- si_phenology_simulate(12, 10, shift_mean = -10, shift_sd = 25, seed = 1)

fl <- si_forbidden(
  phenology  = si_rule_phenology(ph, "current", threshold = 0.3),
  morphology = si_rule_morphology(runif(12, 5, 40), runif(10, 3, 35),
                                  rule = "barrier"),
  binarise   = TRUE)
fl
si_forbidden_partition(fl)
```

`si_rule_custom()` is the extension point: any function of the two species
pools returning a matrix on `[0, 1]` is a valid rule.

## 4. Comparing two time slices

```{r}
A1 <- si_simulate_species(12, 22, 22, prefix = "plant", seed = 2)
B1 <- si_simulate_species(10, 22, 22, prefix = "poll",  seed = 3)
A2 <- si_simulate_climate(A1, "decline", severity = 0.35)
B2 <- si_simulate_climate(B1, "decline", severity = 0.35)

mw2 <- si_simulate_metaweb(12, 10, 0.3, seed = 2,
                           names_A = A1$names, names_B = B1$names)
th  <- function(s) si_threshold(s, "fixed", 0.3)

f1 <- si_overlap(th(A1), th(B1), mw2, method = "binary",
                 scenario = "current", check = FALSE)
f2 <- si_overlap(th(A2), th(B2), mw2, method = "binary",
                 scenario = "future",  check = FALSE)

table(si_network_extinction(f1, f2)$status)
```

Now the partition of interaction turnover:

```{r}
b <- suppressWarnings(si_beta(f1, f2))
round(colMeans(b[, c("beta_S", "beta_WN", "beta_ST", "beta_OS")],
               na.rm = TRUE), 6)
```

`beta_OS` is exactly zero, and `si_beta()` warned you why: both fields share
one constraint layer. Species present at both times are linked at both times or
at neither, so no shared species can change partners. **A static constraint
layer makes rewiring impossible by construction, whatever the climate scenario.**

```{r}
sum(si_rewiring(f1, f2)$realised)   # new links among shared species
```

## 5. Making rewiring possible

Rewiring requires the constraint layer itself to change. The clearest
mechanism is differential phenological shift.

```{r}
gains <- function(shift_sd) {
  ph <- si_phenology_simulate(12, 10, shift_mean = -10,
                              shift_sd = shift_sd, seed = 4)
  m1 <- si_metaweb(si_rule_phenology(ph, "current", 0.3)$matrix,
                   names_A = A1$names, names_B = B1$names)
  m2 <- si_metaweb(si_rule_phenology(ph, "future",  0.3)$matrix,
                   names_A = A1$names, names_B = B1$names)
  g1 <- si_overlap(th(A1), th(B1), m1, method = "binary", check = FALSE)
  g2 <- si_overlap(th(A2), th(B2), m2, method = "binary", check = FALSE)
  c(beta_OS = mean(si_beta(g1, g2)$beta_OS, na.rm = TRUE),
    gains   = sum(si_rewiring(g1, g2)$realised))
}

t(vapply(c(0, 10, 30, 60), gains, numeric(2)))
```

At `shift_sd = 0` every species advances by the same ten days, phenological
overlap is unchanged, and rewiring is again exactly zero. Rewiring is
generated by the *variance* of the shift among species, not by its mean.

## 6. Robustness, with and without constraints

```{r}
w <- mw2$matrix * matrix(runif(120, 1, 10), 12, 10)
rew <- si_rule_morphology(runif(12, 5, 40), runif(10, 3, 35),
                          rule = "gower")$matrix

r0 <- si_robustness(w, remove = "A", n_sim = 60, seed = 1)
r1 <- si_robustness(w, remove = "A", rewiring = matrix(1, 12, 10),
                    n_sim = 60, seed = 1)
r2 <- si_robustness(w, remove = "A", rewiring = rew, n_sim = 60, seed = 1)

round(c(R = r0$robustness, Rw_unconstrained = r1$robustness,
        Rw_trait_constrained = r2$robustness), 4)
```

Unconstrained partner switching is an upper bound that trait matching does not
support.

## 7. Beyond mutualism

The sign structure of the interaction controls the coextinction rule and the
direction in which spatial overlap should be read.

```{r}
si_interaction("eatenBy", level_A = "crop", level_B = "pest",
               overlap_sign = -1)
si_cascade_rule(si_interaction("+-"))
```

For an antagonism the resource level is *released* by the loss of its
consumers rather than coextinct, and increasing future overlap is reported as
a deterioration by `si_compare()` and `si_service_change()`.

## 8. Virtual experiments

```{r}
des <- list(richness = 15, connectance = c(1, 0.5, 0.2),
            severity = 0.4, grid = 12)
ex <- suppressWarnings(si_experiment(des, replicates = 3, cells = 30, seed = 1))
round(aggregate(cbind(d_connectance, d_NODF, beta_WN) ~ connectance,
                ex, mean), 4)
```

Topological responses are exactly zero when the constraint layer is
unconstrained, and grow as constraints tighten. `si_sensitivity()` decomposes
the variance across design factors.

## 9. The bundled Chilean data

```{r}
data(chile_pp)
c(cells = nrow(chile_pp$plants_current),
  plants = ncol(chile_pp$plants_current),
  pollinators = ncol(chile_pp$pollinators_current),
  links = sum(chile_pp$metaweb > 0))
```

```{r}
P  <- si_stack(chile_pp$plants_current_bin,      level = "plant")
A  <- si_stack(chile_pp$pollinators_current_bin, level = "pollinator")
P$template <- A$template <- chile_pp$grid
P$cells    <- A$cells    <- chile_pp$cells

fc <- si_overlap(P, A, si_metaweb(chile_pp$metaweb), method = "binary",
                 check = FALSE)
summary(si_metrics(fc, what = c("links", "connectance", "NODF")))
```

The full 187 × 171 analysis is reproduced by the script in
`system.file("scripts", "chile_case_study.R", package = "spinet")`.
